### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/1-introduction.tex
Finishing revision draft, few more red highlights to add.

The currently available software for monotone piecewise polynomial
interpolation includes quadratic [He and Shi 1998, McAllister and
Roulier 1981], cubic [Fritsch and Carlson 1980], and (with limited
application) quartic [Wang and Tan 2004; Piah and Unsworth 2011; Yao
and Nelson 2018] cases. In addition, a statistical method for
bootstrapping the construction of an arbitrarily smooth monotone fit
exists [Leitenstorfer and Tutz 2006], but the method does not take
advantage of known analytic properties of quintic polynomials.
Conversely the work on quadratic shape preserving splines by Schumaker
[1983] is widely utilized in theory for its conservation of both
monotonicity and convexity, but no mathematical software is published.

The code by Fritsch [1982] for $C^1$ cubic spline interpolation is the
predominantly utilized code for constructing monotone interpolants at
present. Theory has been provided for the quintic case [Ulrich and
Watson 1994; He{\ss} and Schmidt 1994] and that theory was recently
utilized in a proposed algorithm [Lux 2020] for monotone quintic
spline construction, however no published mathematical software
exists. The software presented here represents the first published
software package for producing $C^2$ shape preserving splines. This
work improves upon the algorithm presented by Lux et al. [2020] by
refactoring computations for improved numerical stability, estimating
minimum curvature derivatives at breakpoints with a quadratic facet
model, and using a binary search to reduce the magnitude of the
modifications made to initial derivative estimates when constructing a
monotone spline interpolant.

This work provides a Fortran 2003 subroutine {\tt MQSI} based on the
necessary and sufficient conditions in Ulrich and Watson [1994] for the
construction of monotone quintic spline interpolants of monotone data.
Precisely, the problem is, given a strictly increasing sequence $X_1<X_2<
\cdots <X_n$ of breakpoints with corresponding monotone increasing
function values $Y_1\le Y_2\le \cdots \le Y_n$, find a $C^2$ monotone
increasing quintic spline $Q(x)$ with the same breakpoints satisfying
$Q(X_i)=Y_i$ for $1\le i\le n$. ({\tt MQSI} actually does something slightly
more general, producing $Q(x)$ that is monotone increasing (decreasing)
wherever the data is monotone increasing (decreasing).)

>>>>

The existing research in shape preserving interpolatory splines is
rich and filled with many approaches for many different
applications. In the context of this work, the unique quality of shape
preserving spline algorithms can be observed through: (1) the type of
spline (rational or Hermite), (2) the polynomial order and level of
continuity, and (3) the sharpness, sufficiency, and necessity of the
conditions used to establish shape preserving properties. While the
choice of spline representation is less important, achieving higher
levels of continuity is often more difficult than lower levels of
continuity. And while all referenced algorithms establish sufficient
conditions for shape preservation, it is less common and much more
difficult to provide sharp {\it necessary} conditions for
monotonicity. The following two paragraphs highlight a collection of
research related to the construction of sufficient conditions for
$C^1$ and $C^2$ shape preserving splines. From this foundation, the
value and utility of the code presented here can be more readily
appreciated.

DeVore [1977] proves that monotone spline interpolants have improved
accuracy over their nonmonotone counterparts when approximating
monotone functions, while Constantini [1986] proves that monotone
spline interpolants to data of fixed continuity and order exist in
general. This groundwork establishes the potential for monotone spline
interpolation. Schumaker [1983] and McAllister and Roulier [1981]
(independently) establish necessary and sufficient conditions for
monotone $C^1$ quadratic splines through the insertion of additional
knots and Fritsch [1982] establishes simplified sufficient conditions
for monotone $C^1$ cubic splines that require no knot insertion.
Gregory and Delbourgo [1983] prove necessary and sufficient conditions
for a closed form solution to monotone $C^1$ quadratic rational spline
interpolation with nonlinear boundary equations and an iterative
approach. Delbourgo and Gregory [1993] extend that same work to
achieve sufficient conditions for $C^2$ continuity with a cubic
rational spline given additional tension parameters defined by
users. Huynh [1993] similarly arrives at monotone $C^1$ cubic spline
interpolants and several necessary nonlinear boundary conditions on
monotonicity.

Continuing towards monotone $C^2$ spline interpolation, Fiorot and
Tabka [1991] prove a simple method for determining the existence of a
monotone $C^2$ cubic interpolating spline, but note that functions of
this kind do not always exist for arbitrary convex monotone data sets.
Pruess [1993] proposes sufficient conditions for a cubic $C^2$ shape
preserving spline method while acknowledging that quintic splines are
necessary for generally monotone $C^2$ spline interpolants that do not
insert new knots. Similarly, Manni [1997] constructs sufficient
conditions for a monotone $C^2$ cubic spline by adding two additional
knots per interval, then Cravery and Manni [2003] extend that
methodology to arrive at monotone $C^3$ interpolating splines by
progressively increasing tension through Bezier control
polygons. Constantini [1997] provides a method for monotone $C^2$
quintic spline interpolation based on sufficient and necessary
boundary value conditions, but the necessary conditions are not
sharp. Dougherty, Edelman, and Hyman [1989] construct monotone $C^2$
Hermite splines and in commentary recommend against the direct
optimization for smoothness parameters like global $L^2$, rather
promoting problem specific definitions of geometric niceness. Both
Wang and Tan [2004] as well as Yao and Nelson [2018] propose
sufficient conditions for $C^2$ quartic splines, while Piah and
Unsworth [2011] improve upon those sufficient conditions for $C^2$
quartic rational spline interpolation. Abbas, Majid, and Ali [2012]
formulate a set of sufficient conditions for monotone $C^2$ cubic
rational splines and similar work is extended to $C^2$ quartic
rationals (and splines with arbitrary smoothness) by Zhu and Han
[2015]. Ulrich and Watson [1994] arrived at sufficient and sharp
necessary conditions for monotone $C^2$ quintic splines, as also did
He{\ss} and Schmidt [1994] (independently). In addition, a statistical
method for bootstrapping the construction of an arbitrarily smooth
monotone fit exists [Leitenstorfer and Tutz 2006], but this is enabled
by sufficient conditions only.

The currently available peer reviewed and published software for
monotone piecewise polynomial interpolation is severely reduced in
comparison with the number of published approaches mentioned
above. This is partially indicative of how difficult it is to properly
handle the numerical conditions that arise when constructing and
evaluating precise shape preserving splines, as well as how difficult
it is to create code that correctly conforms with the theoretical
expectations. The sufficient $C^1$ cubic spline method PCHIP of
Fritsch and Carlson [1980] is available through the SciPy Python
package ($>1$ billion downloads), and the $C^1$ quadratic method of
Schumaker [1983] is available as an R package (300 thousand
downloads). The piecewise quadratic $C^1$ method of McAllister and
Roulier [1981] is available in FORTRAN (1000 downloads) and the
sufficient $C^2$ quintic method of Costantini [1997] BVSPIS in FORTRAN
as well (500 downloads). Based on publicly available download records,
it is assumed that the code by Fritsch [1982] for monotone $C^1$ cubic
spline interpolation is the predominant code for constructing monotone
interpolants at present.

The theory for sufficient and {\it sharp} necessary bounds for
monotone quintic interpolation has been provided [Ulrich and Watson
  1994; He{\ss} and Schmidt 1994] and that theory was recently
utilized in a proposed algorithm [Lux 2020] for monotone quintic
spline construction, however no published mathematical software exists
for the quintic case based on sharp monotonicity conditions. The
software presented here represents the first published software
package for producing $C^2$ shape preserving splines based on sharp
monotonicity conditions. This work improves upon the algorithm
presented by Lux et al. [2020] by refactoring computations for
improved numerical stability, estimating minimum magnitude second
derivatives at breakpoints with a quadratic facet model, and using a
binary search to reduce the magnitude of the modifications made to
initial derivative estimates when constructing a monotone spline
interpolant.

\heading{Overview}

This work provides a Fortran 2003 subroutine {\tt MQSI} based on the
sharp necessary and sufficient conditions in Ulrich and Watson [1994]
for the construction of monotone quintic spline interpolants of
monotone data. Precisely, the problem is, given a strictly increasing
sequence $X_1<X_2< \cdots <X_n$ of breakpoints with corresponding
monotone increasing function values $Y_1\le Y_2\le \cdots \le Y_n$,
find a $C^2$ monotone increasing quintic spline $Q(x)$ with the same
breakpoints satisfying $Q(X_i)=Y_i$ for $1\le i\le n$. ({\tt MQSI}
actually does something slightly more general, producing $Q(x)$ that
is monotone increasing (decreasing) wherever the data is monotone
increasing (decreasing).)

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/mqsi20.tex
Finishing revision draft, few more red highlights to add.

@@

@@

@@

% haralick1981facet
\refj{Haralick, R. M., and Watson, L. T.}{1981}{A facet model for
  image data.}{Computer Graphics and Image Processing}{15(2):113--129}

% he1998monotone
\refj{He, X. and Shi, P.}{1998}{Monotone B-spline smoothing}{Journal
  of the American Statistical Association}{93(442):643--650}

@@

@@

@@

@@

@@

>>>>

%% No extra knots, rational functions, sufficiency and necessity
%% proven, works for unequally spaced data, no code associated with
%% the paper, no algorithm provided for the initialization for free
%% shape parameters.
\refj{Abbas M., Majid, A. A. and Ali,
  J. M.}{2012}{Monotonicity-preserving $C^2$ rational cubic spline for
  monotone data}{Applied Mathematics and
  Computation}{219(6):2885--2895}

%% Develops a quadratic spline curve fitting algorithm that maintains
%% local convexity while also proving some error bounds. This method
%% avoids inserting additional knots, but does not interpolate the
%% data.
%% \refj{Beatson, R. K.}{1982}{Monotone and convex approximation by
%%   splines: error estimates and a curve fitting algorithm}{SIAM Journal
%%   on Numerical Analysis}{19(6):1278--1285}

@@

%% Proves existence of splines that maintain monotonicity and
%% convexity while having fixed continuity and order.
\refj{Costantini, P.}{1986}{On monotone and convex spline
  interpolation}{Mathematics of Computation}{46(173):203--214}

%% Source paper accompanying the BVSPIS publication.
\refj{Costantini, P.}{1997}{Boundary-valued shape-preserving
  interpolating splines}{ACM Transactions on Mathematical Software
  (TOMS)}{23(2):229--251}

%% BVSPIS Algorithm.
\refj{Costantini, P.}{1997}{Algorithm 770: BVSPIS—a package for
  computing boundary-valued shape-preserving interpolating
  splines}{ACM Transactions on Mathematical Software (TOMS)}{23(2):252-–254}

%% Constructs C3 interpolating splines by progressively increasing
%% tension through Bezier control polygons until shape parameters are
%% satisfied, no code is provided.
\refj{Cravero, I., C. Manni}{2003}{Shape-preserving interpolants with
  high smoothness}{Journal of computational and applied
  mathematics}{157(2):383--405}

%% Iterates towards a solution of nonlinear equations that ensure C2
%% continuity of quadratic rational splines. The solution they iterate
%% towards exists and is unique. They provide proofs and methods for
%% setting derivatives at the ends, no formal algorithms or code.
\refj{Delbourgo, R., Gregory, J. A.}{1983}{$C^2$ rational quadratic
  spline interpolation to monotonic data}{IMA Journal of Numerical
  Analysis}{3(2):141--152}

%% Continuous previous work by incorporating a cubic numerator, proves
%% shape preservation when the correct value for a tension parameter
%% is set but no method for setting that parameter. Accuracy depends
%% on accurate derivative evaluations. No code nor formal algorithm is
%% provided.
\refj{Delbourgo, R.}{1993}{Accurate $C^2$ rational interpolations in
  tension}{SIAM Journal on Numerical Analysis}{30(2):595--607}

%% Proves approximation error bounds for monotone spline interpolation
%% of data from monotone functions.
\refj{DeVore, R. A.}{1977}{Monotone approximation by splines}{SIAM
  Journal on Mathematical Analysis}{8(5):891--905}

%% Accompanies the previous error bounds in the context of monotone
%% polynomial interpolation of data from a monotone function.
\refj{DeVore, R. A.}{1977}{Monotone approximation by polynomials}{SIAM
  Journal on Mathematical Analysis}{8(5):906--921}

%% Proposes an algorithm for constructing C2 Hermite monotone spline
%% interpolants using proven sufficient, but not necessary
%% conditions. As these bounds are not sharp, the space of allowable
%% derivative and second derivative values at knots is unnecessarily
%% shrunk. An algorithm is provided, but no code. Recommends against
%% the direct optimization for smoothness parameters like L2, instead
%% promoting problem specific definitions of 'geometric niceness'.
\refj{Dougherty, R. L., Edelman, A., Hyman,
  J. M.}{1989}{Nonnegativity-, monotonicity- convexity-preserving
  cubic and quintic Hermite interpolation}{Mathematics of
  Computation}{52(186):471--494}

%% Proves a simple method for determining the existence of a monotone
%% C2 cubic interpolating spline, but functions of this kind do not
%% always exist for arbitrary convex monotone data sets.
\refj{Fiorot, J. C., Tabka, J.}{1991}{Shape-preserving $C^2$ cubic
  polynomial interpolating splines}{Mathematics of
  Computation}{57(195):291--298}

@@

%% Proves necessary and sufficient conditions for a closed form
%% solution to monotone C1 quadratic rational spline
%% interpolation. Does not provide formal algorithm, nor code.
\refj{Gregory , J. A.,Delbourgo, R.}{1982}{Piecewise rational
  quadratic interpolation to monotonic data}{IMA Journal of Numerical
  Analysis}{2(2):123--130}

@@

%% Proposes a method for making arbitrarily smooth shape preserving
%% spline interpolants, but uses a centered quadratic to estimate
%% derivatives and a sufficient set of bounding conditions to ensure
%% shape preservation in practice. No necessary conditions are
%% applied, no algorithm is provided, neither is any code provided.
\refj{Han, X.}{2018}{Shape-preserving piecewise rational interpolation
  with higher order continuity}{Applied Mathematics and
  Computation}{337:1--13}

@@

%% Presents several nonlinear boundary conditions based on the median
%% function to establish a fourth order monotonicity preserving
%% constraints relying on divided difference initial derivative
%% estimates. Overall generates a monotone C1 cubic interpolant to
%% data.
\refj{Huynh, H. T.}{1993}{Accurate monotone cubic interpolation}{SIAM
  Journal of Numerical Analysis}{30(1):57--100}

@@

%% Use sufficient conditions for monotonicity and convexity to form a
%% globally weighted combination of C1 quadratic splines that satisfy
%% shape constraints.
%% \refj{Kvasov, B. I.}{2014}{Monotone and convex interpolation by
%%   weighted quadratic splines}{Advances in Computational
%%   Mathematics}{40(1):91--116}

@@

%% Constructs a C2 cubic spline with two additional knots per
%% interval. No code (paywall).
\refj{Manni, C., Sablonni\`ere, P.}{1997}{Monotone interpolation of
  order 3 by $C^2$ cubic splines}{IMA Journal of Numerical
  Analysis}{17(2):305--320}

%% Uses sufficient conditions to generate C2 cubic spline. No code
%% (paywall).
\refj{Manni, C.}{2001}{On shape preserving $C^2$ Hermite
  interpolation}{BIT Numerical Mathematics}{41(1):127--148}

@@

%% Proposes a cubic C2 shape preserving spline method with free
%% parameters (mentions quintic splines are necessary for globally C2
%% monotone splines). Sufficient (but not necessary) conditions are
%% provided for monotonicity.
\refj{Pruess, S.}{1993}{Shape preserving $C^2$ cubic spline
  interpolation}{IMA Journal of Numerical Analysis}{13(4):493--507}

@@

%% Uses sufficient conditions over three free parameters to construct
%% C2 rational quartic shape preserving splines.
\refj{Zhu, Y.,Han, X.}{2015}{$C^2$ rational quartic interpolation
  spline with local shape preserving property}{Applied Mathematics
  Letters}{46:57--63}

%% Uses sufficient conditions over two free parameters to construct C2
%% rational quartic shape preserving splines.
\refj{Zhu, Y.,Han, X.}{2015}{Shape preserving $C^2$ rational quartic
  interpolation spline with two parameters}{International Journal of
  Computer Mathematics}{92(10):2160--2177}
